---
title: brunnermunzel
author: Toshiaki Ara
date: "`r Sys.Date()`"
---

# Introduction

For Brunner-Munzel test
`brunner.munzel.test` function in `lawstat` package is well-known.
`brunnermunzel.test` in this package also accepts **formula**.

Moreover,
 this package provides
 `brunnermunzel.permutation.test` function
  for a permuted Brunner-Munzel test
  used in the case of small sample size.

The script of `brunnermunzel.test` is derived from
 that of `lawstat` package, and is rewrote with FORTRAN.
Thanks to Vyacheslav Lyubchich
 (maintainer of `lawstat` package).

# Installation
```{r, eval=FALSE}
remotes::install_github("toshi-ara/brunnermunzel")
```

# Example
## Brunner-Munzel test
```{r, message=FALSE}
## Pain score on the third day after surgery for 14 patients under
## the treatment Y and 11 patients under the treatment N
## (see Brunner and Munzel, 2000; Neubert and Brunner, 2007).

Y <- c(1, 2, 1, 1, 1, 1, 1, 1, 1, 1, 2, 4, 1, 1)
N <- c(3, 3, 4, 3, 1, 2, 3, 1, 1, 5, 4)

dat <- data.frame(
    value = c(Y, N),
    group = factor(rep(c("Y", "N"), c(length(Y), length(N))),
                   levels = c("Y", "N"))
)

library(brunnermunzel)

## Default
brunnermunzel.test(Y, N)

## Formula interface
brunnermunzel.test(value ~ group, data = dat)
```

## permuted Brunner-Munzel test

`brunnermunzel.permutation.test` takes time to obtain results.

```{r, message=FALSE}
library(brunnermunzel)

Y <- c(1, 2, 1, 1, 1, 1, 1, 1, 1, 1, 2, 4, 1, 1)
N <- c(3, 3, 4, 3, 1, 2, 3, 1, 1, 5, 4)

dat <- data.frame(
    value = c(Y, N),
    group = factor(rep(c("Y", "N"), c(length(Y), length(N))),
                   levels = c("Y", "N"))
)

## Default
brunnermunzel.permutation.test(Y, N)

## Formula interface
brunnermunzel.permutation.test(value ~ group, data = dat)
```

### Comparison with other methods
```{r, message=FALSE}
# Wilcoxon sum-rank test
wilcox.test(value ~ group, data = dat)

# exact Wilcoxon sum-rank test
library(coin)
wilcox_test(value ~ group, data = dat, distribution = "exact")
```

