---
title: "Usage of `brunnermunzel` package"
author: "Toshiaki Ara"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Usage of brunnermunzel package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This `brunnermunzel` package is to perform (permuted) Brunner-Munzel test
 for stochastic equality of two samples,
 which is also known as the Generalized Wilcoxon test.

For **Brunner-Munzel test** [@ref:1],
`brunner.munzel.test` function in `lawstat` package is very famous.
This function is extended to enable
 to use **formula**, **matrix**, and **table** as an argument.

Also, the function `brunnermunzel.permutation.test`
 for **permuted Brunner-Munzel test** [@ref:2] was provided.

# Usage of functions in `brunnermunzel` package
## Default and Formula class
### Sample data

In this section,
 we will use sample data from Hollander & Wolfe (1973), 29f.
 -- Hamilton depression scale factor measurements in 9 patients with
 mixed anxiety and depression, taken at the first (x) and second (y) visit
 after initiation of a therapy (administration of a tranquilizer)".

```{r}
x <- c(1.83,  0.50,  1.62,  2.48, 1.68, 1.88, 1.55, 3.06, 1.30)
y <- c(0.878, 0.647, 0.598, 2.05, 1.06, 1.29, 1.06, 3.14, 1.29)
```

For formula interface, data was converted to `data.frame`.

```{r}
dat <- data.frame(
    value = c(x, y),
    group = factor(rep(c("x", "y"), c(length(x), length(y))),
                   levels = c("x", "y")))
```

```{r, message=FALSE}
library(dplyr)
dat %>%
    group_by(group) %>%
    summarize_all(list(mean = mean, median = median))
```


```{r, echo=FALSE, fig.width=4, fig.asp=1, out.width="50%"}
library(ggplot2)
set.seed(100)
ggplot(dat, aes(x = group, y = value)) +
    geom_jitter(width = 0.01)
```

### Analysis
#### analysis with Brunner-Munzel test
```{r}
library(brunnermunzel)

brunnermunzel.test(x, y)

brunnermunzel.test(value ~ group, data = dat)
```

#### analysis with permuted Brunner-Munzel test
To perform permuted Brunner-Munzel test,
 use `brunnermunzel.test` with "`perm = TRUE`" option,
 or `brunnermunzel.permutation.test` function.
This "`perm`" option is used in also formula interface, matrix, and table.

When `perm` is `TRUE`,
 `brunnermunzel.test` calls
 `brunnermunzel.permutation.test` in internal.

```{r, message=FALSE}
brunnermunzel.test(x, y, perm = TRUE)

brunnermunzel.permutation.test(x, y)
```

Because statistics in all combinations are calculated
 in permuted Brunner-Munzel test
 (${}_{n_{x}+n_{y}}C_{n_{x}}$ where
  $n_{x}$ and $n_{y}$ are sample size of $x$ and $y$, respectively),
  it takes a long time to obtain results.

Therefore, when sample size is too large
 [the number of combination is more than 40116600
  ($=$ `choose(28, 14)`)],
 it switches to Brunner-Munzel test automatically.

```{r}
# sample size is 30
brunnermunzel.permutation.test(1:15, 3:17)
```

#### using `force` option

When you want to perform permuted Brunner-Munzel test
 regardless sample size,
 you add "`force = TRUE`" option to `brunnermunzel.permutation test`.

```{r, eval=FALSE}
brunnermunzel.permutation.test(1:15, 3:17, force = TRUE)
#>
#>  permuted Brunner-Munzel Test
#>
#> data:  1:15 and 3:17
#> p-value = 0.2341
```

#### using `alternative` option

`brunnermunzel.test` also can use "`alternative`" option
 as well as `t.test` and `wilcox.test` functions.

To test whether the average rank of group $x$ is greater
 than that of group $y$,
 `alternative = "greater"` option is added.
In contrast,
 to test whether the average rank of group $x$ is lesser
 than that of group $y$,
 `alternative = "less"` option is added.

The results of Brunner-Munzel test and
 Wilcoxon sum-rank test (Mann-Whitney test)
 with `alternative = "greater"` option are shown.
In this case,
 median of $x$ is `r round(median(x), 2)`, and
 median of $y$ is `r round(median(y), 2)`.

```{r}
brunnermunzel.test(x, y, alternative = "greater")

wilcox.test(x, y, alternative = "greater")
```

When using formula,
 `brunnermunzel.test` with `alternative = "greater"` option
 tests an alternative hypothesis "1st level is greater than 2nd level".

In contrast,
 `brunnermunzel.test` with `alternative = "less"` option
 tests an alternative hypothesis "1st level is lesser than 2nd level".

```{r}
dat$group
```

```{r}
brunnermunzel.test(value ~ group, data = dat, alternative = "greater")$p.value

wilcox.test(value ~ group, data = dat, alternative = "greater")$p.value
```

```{r}
brunnermunzel.test(x, y, alternative = "less")$p.value

wilcox.test(x, y, alternative = "less")$p.value
```

#### using `est` option

Normally, `brunnermunzel.test` and `brunnermunzel.permutation test`
 return the estimate $P(X<Y) + 0.5 \times P(X=Y)$.
When '`est = "difference"`' option is used,
 these functions return mean difference [$P(X<Y) - P(X>Y)$]
  in estimate and confidence interval.

Note that $P(X<Y) - P(X>Y) = 2p - 1$
 when $p = P(X<Y) + 0.5 \times P(X=Y)$.

This change is proposed by Dr. Julian D. Karch.


```{r}
brunnermunzel.test(x, y, est = "difference")

brunnermunzel.permutation.test(x, y, est = "difference")
```



## Matrix and Table class

In some case, data is provided as aggregated table.
Both `brunnermunzel.test` and `brunnermunzel.permutation.test`
 accept data of **matirix** and **table** class.

```{r, sample_data_matrix, echo=FALSE}
dat1 <- matrix(c(5, 3, 2, 1, 3, 6), nr = 2, byrow = TRUE)
dat2 <- as.table(dat1)
colnames(dat2) <- c("Normal", "Moderate", "Severe")
```

```{r, echo=FALSE}
knitr::kable(dat2, caption = "Fictional data")
```

### Sample data

```{r, sample_data_matrix, eval=FALSE}
```

```{r}
dat1  # matrix class

dat2  # table class
```

### Analysis
#### analysis with Brunner-Munzel test

```{r}
brunnermunzel.test(dat1)

brunnermunzel.test(dat2)
```

#### analysis with permuted Brunner-Munzel test

```{r}
brunnermunzel.permutation.test(dat1)

brunnermunzel.permutation.test(dat2)
```

# About program
## `brunnermunzel.test` function

`brunnermunzel.test` function is derived from
 `brunner.munzel.test` function in `lawstat` package
 (Maintainer of this package is Vyacheslav Lyubchich;
  License is GPL-2 or GPL-3)
 with modification.
The authors of this function are
 Wallace Hui, Yulia R. Gel, Joseph L. Gastwirth and Weiwen Miao.

## `combination` subroutine by FORTRAN77

FORTRAN subroutine `combination` in combination.f is derived from
 the program by shikino
 (http://slpr.sakura.ne.jp/qp/combination)(CC-BY-4.0)
  with slight modification.

Without this subroutine,
 I could not make `brunnermunzel.permutation.test`.
Thanks to shikono for your useful subroutine.


# References
